Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-02 EconGrowth.Rnw
Trying to make a CD predictions column that accounts for energy, but failing so far.

<<cobb-douglas, echo=FALSE>>=
####################
# Returns an nls Cobb-Douglas model for the country specified
##
cobbDouglasModel <- function(countryName){
  dataTable <- loadData(countryName) #Load the data that we need.
  # Establish guess values for alpha and lambda.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.7 #0.7 gives good results for all countries.
  # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
  modelCD <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha), 
                 data=dataTable, 
                 start = list(lambda=lambdaGuess, alpha=alphaGuess),
                 algorithm = "port",
                 lower = list(lambda=-Inf, alpha=0), #constraints on lambda and alpha
                 upper = list(lambda=Inf, alpha=1)
                 )
  return(modelCD)
}

#########################
# Takes the Cobb-Douglas fitted models and creates per-country predictions for them.
# Returns a data.frame with the predictions.
##
cobbDouglasPredictions <- function(countryName){
  model <- cobbDouglasModel(countryName)
  pred <- predict(model) #See http://stackoverflow.com/questions/9918807/how-get-plot-from-nls-in-r
  df <- data.frame(pred)
  # There are two countries which do not have labor data for 2011. Need to add a "NA" for 2011 for those countries.
  if (countryName == "CN" || countryName == "ZA"){
    dfNA <- data.frame(NA)
    colnames(dfNA) <- "pred" #Need to give it the same name so that we can apply rbind
    df <- rbind(df, dfNA)
  }
  return(df)
}

#########################
# Takes the Cobb-Douglas fitted models and creates a single column of predicted GDP values
# that corresponds, row for row, with the AllData.txt file.
##
cobbDouglasPredictionsColumn <- function(){
  out <- do.call("rbind", lapply(countryAbbrevs, cobbDouglasPredictions)) #rbinds all the predictions together
  colnames(out) <- c("predGDP")
  return(out)
}

>>>>

<<cobb-douglas, echo=FALSE>>=
####################
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
##
cobbDouglasModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.

  betaGuess <- 0.3 #if needed
  #Run the non-linear least squares fit to the data.
  if (missing(energyType)){
    # No energy term
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    modelCD <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha), 
                   data=dataTable, 
                   algorithm = "port",
                   start = list(lambda=lambdaGuess, alpha=alphaGuess),
                   lower = list(lambda=-Inf, alpha=0), #constraints on lambda and alpha
                   upper = list(lambda=Inf, alpha=1))
    return(modelCD)
  } else {
    energyColumnName <- paste("i", energyType, sep="")
    iEnergy <- dataTable[energyColumnName] #grabs the desired energy column
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.4 
    betaGuess <- 1.0 - alphaGuess
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess - 0.05
    modelCDe <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEnergy^(1.0 - max(a,b)), 
                    data = dataTable,
                    algorithm = "port",
                    start = list(lambda=lambdaGuess, a=aGuess, b=bGuess),
                    lower = list(lambda=-Inf, a=0, b=0),
                    upper = list(lambda= Inf, a=1, b=1),
                    control = nls.control(maxiter = 200, 
                                          tol = 1e-05, 
                                          minFactor = 1/1024, 
                                          printEval=TRUE, 
                                          warnOnly=TRUE))
    return(modelCDe)
  }
}
#print(lapply(countryAbbrevs, cobbDouglasModel, energyType="Q"))
model <- cobbDouglasModel(countryAbbrev="ZM", energyType="Q")
print(model)
print(model$convInfo)
#print(lapply(countryAbbrevs, cobbDouglasModel))

#########################
# Takes the Cobb-Douglas fitted models and creates per-country predictions for them.
# Returns a data.frame with the predictions.
##
cobbDouglasPredictions <- function(countryAbbrev, energyType){
  model <- cobbDouglasModel(countryAbbrev, energyType)
  pred <- predict(model) #See http://stackoverflow.com/questions/9918807/how-get-plot-from-nls-in-r
  df <- data.frame(pred)
  # There are two countries which do not have labor data for 2011. Need to add a "NA" for 2011 for those countries.
  if (countryAbbrev == "CN" || countryAbbrev == "ZA"){
    dfNA <- data.frame(NA)
    colnames(dfNA) <- "pred" #Need to give it the same name so that we can apply rbind
    df <- rbind(df, dfNA)
  }
  return(df)
}

#########################
# Takes the Cobb-Douglas fitted models and creates a single column of predicted GDP values
# that corresponds, row for row, with the AllData.txt file.
##
cobbDouglasPredictionsColumn <- function(energyType){
  if (missing(energyType)){
    out <- do.call("rbind", lapply(countryAbbrevs, cobbDouglasPredictions)) #rbinds all the predictions together
  } else {
    out <- do.call("rbind", lapply(countryAbbrevs, cobbDouglasPredictions, energyType=energyType))
  }
  colnames(out) <- c("predGDP")
  return(out)
}
